Cell Decompositions for Rank Two Quiver Grassmannians

### 2018-02-06 Grassmannian_cells.tex
Section 4 edits

\subsection{Torus Actions}
\label{torusaction}
Fix a vector space $X$ of dimension $n$ and let $k\leq n$.
We first consider a natural torus action on usual Grassmannians coming along with any direct sum decomposition of the vector space $X$.
Then we generalize the concept to quiver Grassmannians and observe that the torus fixed point sets can be calculated in an analogous manner.

Assume that $X=\bigoplus_{j=1}^l X_j$ is a direct sum decomposition of $X$.
We choose a basis $\mathcal B=\{v_1,\ldots,v_n\}$ of~$X$ which is compatible with this decomposition, i.e.\ there exist numbers $0=r_0<r_1<r_2<\cdots<r_{l-1}<r_l=n$ such that
\[v_{r_0+1},\ldots,v_{r_1}\in X_1,\quad v_{r_1+1},\ldots,v_{r_2}\in X_2,\quad\cdots\quad v_{r_{l-1}+1},\ldots,v_{r_l}\in X_l.\]
Moreover, we choose a map $d:\{1,\ldots,n\}\to\ZZ$ such that $d(i)=d(i')$ if $v_i,v_{i'}\in X_j$ for some $j$.
Then there exists a $\C^\ast$-action on $X$ when linearly extending the definition $t.v_i:=t^{d(i)}v_i$ for $i=1,\ldots,n$ to all of~$X$.
This naturally induces an action of $\C^\ast$ on the usual Grassmannians $\Gr_k(X)$.
We can represent each subspace~$U\in\Gr_k(X)$ uniquely as a matrix $M(U)$ in row-echelon form with respect to the basis $\mathcal B$, i.e.\ there exists a unique sequence $1\leq i_1<\ldots<i_k\leq n$ so that $M(U)$ is of the form
\[M(U):=
  \begin{pmatrix}
    \ast&\cdots &\ast &1&0 &\cdots& 0 & 0 & 0 &\cdots&0&0&0&\cdots&0\\
    \ast&\cdots &\ast&0&\ast&\cdots &\ast&1&0&\cdots&0&0&0&\cdots&0\\
    \ast &\cdots&\ast&0&\ast &\cdots&\ast&0&\ast&\cdots&0&0&0&\cdots&0\\[-0.4em]
    \vdots &\ddots&\vdots&\vdots&\vdots &\ddots&\vdots&\vdots&\vdots&\ddots&\vdots&\vdots&\vdots&\ddots&\vdots\\
    \ast&\cdots &\ast&0&\ast&\cdots &\ast&0&\ast&\cdots&0&0&0&\cdots&0\\
    \ast&\cdots &\ast&0&\ast&\cdots &\ast&0&\ast&\cdots&\ast&1&0&\cdots&0
  \end{pmatrix}\in M_{k,n}(k),\]
where the unit vectors are in the columns $\bfi=(i_1,\ldots,i_k)$.
The set of all subspaces of $X$ represented by matrices of this fixed form define the \emph{Schubert cell} $X_\bfi$.
The $\C^\ast$-action is then given by multiplying the $w^{\mathrm{th}}$ column with $t^{d(w)}$, i.e.\ we have
\[(t.M(U))_{u,w}=t^{d(w)}\cdot M(U)_{u,w}\]
for $w=1,\ldots,n$.
Every Schubert cell is invariant under the torus action, i.e.\ $t.M(U)\in X_\bfi$ for all $M(U)\in X_{\bfi}$. 
\begin{lemma}
  \label{usualGrass}
  Consider $d:\{1,\ldots,n\}\to\ZZ$ for which $d(i)\neq d(i')$ if $v_i\in X_j$ and $v_{i'}\in X_{j'}$ with $j\neq j'$.
  Then we have $U\in X_{\bfi}^{\C^\ast}$ if and only if $U=\bigoplus_{j=1}^l U\cap X_j$.
\end{lemma}
\begin{proof}
  If $U=\bigoplus_{j=1}^l U\cap X_j$, we have 
  \[t.U=\bigoplus_{j=1}^l t.(U\cap X_j)=\bigoplus_{j=1}^l t^{d(r_j)}(U\cap X_j)=U.\]
  Assume that $U$ is a torus fixed with $M(U)\in X_{\bfi}$.
  Fix the $w^{\mathrm{th}}$ column $M(U)_w$ and assume that $v_w\in X_j$. 
  As the matrix representation $M(U)$ is uniquely determined, the assumptions on $d$ imply that all matrix coefficients corresponding to a basis element $v_i$ with $v_i\notin X_j$ are zero.
  This means that $\widehat{M(U)}_w\in X_j$, where~$\widehat{M(U)}_w$ is the vector of $U$ corresponding to $M(U)_w$.
  As the colums are linearly independent, this already shows the claim.
\end{proof}
The next step is to generalize this to quiver Grassmannians.
Let $Q$ be an acyclic quiver.
We choose a map $d:\hat Q_0\to\ZZ$ and fix a representation $X\in\rep Q$ which can be lifted to $\hat Q$.
We can consider the decomposition $X_i=\bigoplus_{\chi\in A_Q} X_{(i,\chi)}$ and define a torus action on each $X_{(i,\chi)}$ via $t.x_{(i,\chi)}=t^{d(i,\chi)}x_{(i,\chi)}$ which can be extended linearly to each $X_i$.
Associated to each subspace $U_i$, there is a corresponding subspace $t.U_i$ for each $t\in\CC^*$.
In general, this does not induce a torus action on the quiver Grassmannians $\Gr_\bfe(X)$ since $t.U=(t.U_i)_{i\in Q_0}$ is not necessarily a subrepresentation of $X$ for every $U\in\Gr_\bfe(X)$.
Indeed, for this such an action must satisfy $X_\alpha(t.U_i)\subset t.U_j$ for every arrow $\alpha:i\to j$ of $Q$. 
\begin{lemma}
  Let $X$ be a representation of $Q$ which can be lifted to $\hat Q$.
  Fix an integer $c_\alpha\in\ZZ$ for each~$\alpha\in Q_1$.
  Suppose $d:\hat Q_0\to\ZZ$ satisfies $d(j,\chi+e_\alpha)-d(i,\chi)=c_\alpha$ for each arrow $\alpha:i\to j$ of $Q$ and each $\chi\in A_Q$.
  Then the torus action on $X$ determined by $d$ induces a torus action on $\Gr_\bfe(X)$.
\end{lemma}
\begin{proof} 
  Since $X$ can be lifted, for any arrow $\alpha\in Q_1$ we can write $X_\alpha:X_i\to X_j$ as a block matrix consisting of linear maps $X_{(\alpha,\chi)}:X_{(i,\chi)}\to X_{(j,\chi+e_\alpha)}$ for $\chi\in A_Q$.
  Fix $U\in\Gr_\bfe(X)$ and consider $u_i\in U_i$.
  Then we can write $u_i=\sum_{\chi\in A_Q} u_{(i,\chi)}$ for some vectors $u_{(i,\chi)}\in U_i\cap X_{(i,\chi)}$.
  But observe that for an arrow $\alpha:i\to j$ of $Q$, we have $X_\alpha(u_{(i,\chi)})=X_{(\alpha,\chi)}(u_{(i,\chi)})\in U_j\cap X_{(j,\chi+e_\alpha)}$, say $X_{(\alpha,\chi)}(u_{(i,\chi)})=u_{(j,\chi+e_\alpha)}$.
  It follows that
  \begin{align*}
    X_\alpha(t.u_i)=\sum\limits_{\chi\in A_Q}X_{(\alpha,\chi)}(t.u_{(i,\chi)})=\sum\limits_{\chi\in A_Q} t^{d(i,\chi)}X_{(\alpha,\chi)}(u_{(i,\chi)})=\sum\limits_{\chi\in A_Q} t^{d(i,\chi)}u_{(j,\chi+e_\alpha)}=t^{-c_\alpha}\sum\limits_{\chi\in A_Q} t.u_{(j,\chi+e_\alpha)}
  \end{align*}
  and thus $X_\alpha(t.U_i)\subset t.U_j$ for every arrow $\alpha:i\to j$ of $Q$.
\end{proof}

\begin{lemma}
  \label{degreecondition} 
  Let $X$ be an indecomposable representation of $\hat Q$.
  There exists $d:\mathrm{supp}(X)\to\ZZ$ and $c_\alpha\in\mathbb N_+$ for each $\alpha\in Q_1$ such that
  \begin{enumerate}
    \item for $(i,\chi),\,(i,\chi')\in\mathrm{supp}(X)$ with $\chi\neq\chi'$, we have $d(i,\chi)\ne d(i,\chi')$;
    \item for $(i,\chi),\,(j,\chi')\in\mathrm{supp}(X)$, we have $d(j,\chi')-d(i,\chi)=c_\alpha$ if and only if $\chi'=\chi+e_\alpha$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  For convenience we introduce the notation $Q_1=\{\alpha_1,\ldots,\alpha_\}$.
  Since $X$ is finite-dimensional and indecomposable, the support quiver $\mathrm{supp}(X)$ is a connected and finite subquiver of $\hat Q$.
  In order to prove the statement, we may assume that $(i,0)\in\mathrm{supp}(X)$ for some $i\in Q_0$.
  Let $K$ be the maximal length of a path in $\mathrm{supp}(X)$ starting or ending in $(i,0)$ such that the underlying graph of the path has no cycles.
  This implies that, for $(i',\chi)\in\mathrm{supp}(X)$ with $\chi=\sum_{l=1}^n \kappa_l e_{\alpha_l}$, we have $|\kappa_l|\leq K$.

>>>>

\subsection{Torus Actions}
\label{torusaction}
Fix a vector space $X$ of dimension $n$ and let $k\leq n$.
We first consider a natural torus action on the usual Grassmannian $\Gr_k(X)$ which is compatible with a given direct sum decomposition of the vector space $X$.
Then we generalize the concept to quiver Grassmannians and observe that the torus fixed point sets can be calculated in an analogous manner.

Given a basis $\mathcal B=\{v_1,\ldots,v_n\}$ of $X$ and a map $d:\{1,\ldots,n\}\to\ZZ$, we get a $\C^\ast$-action on $X$ when linearly extending the definition $t.v_r:=t^{d(r)}v_r$ for $r=1,\ldots,n$ to all of $X$.
This naturally induces an action of $\C^\ast$ on the Grassmannian $\Gr_k(X)$.
Our goal is to understand the fixed points of this action.

For this recall that we can represent each subspace~$U\in\Gr_k(X)$ uniquely by a $k\times n$ matrix $M(U)$ whose rows provide a basis for $U$ when expanded as coefficient vectors in the basis $\mathcal B$.
The uniqueness of $M(U)$ comes from requiring that it be in row-echelon form, i.e.\ there exists a unique sequence $1\leq i_1<\ldots<i_k\leq n$ so that $M(U)$ is of the form
\[M(U):=
  \begin{pmatrix}
    \ast&\cdots &\ast &1&0 &\cdots& 0 & 0 & 0 &\cdots&0&0&0&\cdots&0\\
    \ast&\cdots &\ast&0&\ast&\cdots &\ast&1&0&\cdots&0&0&0&\cdots&0\\
    \ast &\cdots&\ast&0&\ast &\cdots&\ast&0&\ast&\cdots&0&0&0&\cdots&0\\[-0.4em]
    \vdots &\ddots&\vdots&\vdots&\vdots &\ddots&\vdots&\vdots&\vdots&\ddots&\vdots&\vdots&\vdots&\ddots&\vdots\\
    \ast&\cdots &\ast&0&\ast&\cdots &\ast&0&\ast&\cdots&0&0&0&\cdots&0\\
    \ast&\cdots &\ast&0&\ast&\cdots &\ast&0&\ast&\cdots&\ast&1&0&\cdots&0
  \end{pmatrix}\in M_{k,n}(\CC),\]
where the unit vectors are in the columns $\bfi=(i_1,\ldots,i_k)$.
The set of all $U\in\Gr_k(X)$ represented by matrices of this fixed form gives the \emph{Schubert cell} $X_\bfi$.

The $\C^\ast$-action on $U\in\Gr_k(X)$ can then be described in terms of the matrix representation $M(U)$, that is for $U\in X_\bfi$ we have
\[M(t.U)_{qr}=t^{d(r)-d(i_q)} M(U)_{qr}\]
for $q=1,\ldots,k$ and $r=1,\ldots,n$.
Observe that each Schubert cell $X_\bfi$ is invariant under this torus action.

Assume that $X=\bigoplus_{l=1}^m X_l$ is a direct sum decomposition of $X$ and fix a basis $\mathcal B=\{v_1,\ldots,v_n\}$ of~$X$ which is compatible with this decomposition, i.e.\ there exist numbers $0=r_0<r_1<r_2<\cdots<r_{m-1}<r_m=n$ such that
\[v_{r_0+1},\ldots,v_{r_1}\in X_1,\quad v_{r_1+1},\ldots,v_{r_2}\in X_2,\quad\cdots\quad v_{r_{m-1}+1},\ldots,v_{r_m}\in X_m.\]
\begin{lemma}
  \label{usualGrass}
  Consider a map $d:\{1,\ldots,n\}\to\ZZ$ such that $d(r)=d(r')$ if $v_r,v_{r'}\in X_l$ for some $l$ and $d(r)\neq d(r')$ if $v_r\in X_l$ and $v_{r'}\in X_{l'}$ with $l\neq l'$.
  Then under the $\CC^*$-action determined by $d$, we have $U\in\Gr_k(X)^{\C^\ast}$ if and only if $U=\bigoplus_{l=1}^m U\cap X_l$.
\end{lemma}
\begin{proof}
  Assume $U=\bigoplus_{l=1}^m U\cap X_l$.
  Then any $u\in U$ can be written uniquely as $u=\sum_{l=1}^m u_l$ for some $u_l\in U\cap X_l$.
  It follows that $t.u=\sum_{l=1}^m t.u_l=\sum_{l=1}^m t^{d(r_l)}u_l\in\bigoplus_{l=1}^m U\cap X_l=U$ and thus $t.U=U$.

  For the reverse direction, assume $U\in X_{\bfi}$ is a torus fixed point represented by the matrix $M(U)$.
  Then, if~$v_{i_q}\in X_l$, the assumptions on $d$ imply $M(U)_{qr}=0$ unless $r_{l-1}+1\le r\le r_l$.
  That is, $M(U)$ has the shape of a block matrix representing the decomposition $U=\bigoplus_{l=1}^m U\cap X_l$.
\end{proof}
The next step is to generalize this to quiver Grassmannians.
Let $Q$ be an acyclic quiver.
We choose a map $d:\hat Q_0\to\ZZ$ and fix a representation $X\in\rep Q$ which can be lifted to $\hat Q$.
We can consider the decomposition $X_i=\bigoplus_{\chi\in A_Q} X_{(i,\chi)}$ and define a torus action on each $X_{(i,\chi)}$ via $t.x_{(i,\chi)}=t^{d(i,\chi)}x_{(i,\chi)}$ which can be extended linearly to each $X_i$.
Associated to each subspace $U_i$, there is a corresponding subspace $t.U_i$ for each $t\in\CC^*$.
In general, this does not induce a torus action on the quiver Grassmannians $\Gr_\bfe(X)$ since $t.U=(t.U_i)_{i\in Q_0}$ is not necessarily a subrepresentation of $X$ for every $U\in\Gr_\bfe(X)$.
Indeed, for this such an action must satisfy $X_\alpha(t.U_i)\subset t.U_j$ for every arrow $\alpha:i\to j$ of $Q$. 
\begin{lemma}
  Let $X$ be a representation of $Q$ which can be lifted to $\hat Q$.
  Fix an integer $c_\alpha\in\ZZ$ for each~$\alpha\in Q_1$.
  Suppose $d:\hat Q_0\to\ZZ$ satisfies $d(j,\chi+e_\alpha)-d(i,\chi)=c_\alpha$ for each arrow $\alpha:i\to j$ of $Q$ and each $\chi\in A_Q$.
  Then the torus action on $X$ determined by $d$ induces a torus action on $\Gr_\bfe(X)$.
\end{lemma}
\begin{proof} 
  Since $X$ can be lifted, for any arrow $\alpha\in Q_1$ we can write $X_\alpha:X_i\to X_j$ as a block matrix consisting of linear maps $X_{(\alpha,\chi)}:X_{(i,\chi)}\to X_{(j,\chi+e_\alpha)}$ for $\chi\in A_Q$.
  Fix $U\in\Gr_\bfe(X)$ and consider $u_i\in U_i$.
  Then we can write $u_i=\sum_{\chi\in A_Q} u_{(i,\chi)}$ for some vectors $u_{(i,\chi)}\in U_i\cap X_{(i,\chi)}$.
  But observe that for an arrow $\alpha:i\to j$ of $Q$, we have $X_\alpha(u_{(i,\chi)})=X_{(\alpha,\chi)}(u_{(i,\chi)})\in U_j\cap X_{(j,\chi+e_\alpha)}$, say $X_{(\alpha,\chi)}(u_{(i,\chi)})=u_{(j,\chi+e_\alpha)}$.
  It follows that
  \begin{align*}
    X_\alpha(t.u_i)=\sum\limits_{\chi\in A_Q}X_{(\alpha,\chi)}(t.u_{(i,\chi)})=\sum\limits_{\chi\in A_Q} t^{d(i,\chi)}X_{(\alpha,\chi)}(u_{(i,\chi)})=\sum\limits_{\chi\in A_Q} t^{d(i,\chi)}u_{(j,\chi+e_\alpha)}=t^{-c_\alpha}\sum\limits_{\chi\in A_Q} t.u_{(j,\chi+e_\alpha)}
  \end{align*}
  and thus $X_\alpha(t.U_i)\subset t.U_j$ for every arrow $\alpha:i\to j$ of $Q$.
\end{proof}

\begin{lemma}
  \label{degreecondition} 
  Let $X$ be an indecomposable representation of $\hat Q$.
  There exists $d:\mathrm{supp}(X)\to\ZZ$ and $c_\alpha\in\mathbb N_+$ for each $\alpha\in Q_1$ such that
  \begin{enumerate}
    \item for $(i,\chi),\,(i,\chi')\in\mathrm{supp}(X)$ with $\chi\neq\chi'$, we have $d(i,\chi)\ne d(i,\chi')$;
    \item for $(i,\chi),\,(j,\chi')\in\mathrm{supp}(X)$, we have $d(j,\chi')-d(i,\chi)=c_\alpha$ if and only if $\chi'=\chi+e_\alpha$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  For convenience we introduce the notation $Q_1=\{\alpha_1,\ldots,\alpha_n\}$.
  Since $X$ is finite-dimensional and indecomposable, the support quiver $\mathrm{supp}(X)$ is a connected and finite subquiver of $\hat Q$.
  In order to prove the statement, we may assume that $(i,0)\in\mathrm{supp}(X)$ for some $i\in Q_0$.
  Let $K$ be the maximal length of a path in $\mathrm{supp}(X)$ starting or ending in $(i,0)$ such that the underlying graph of the path has no cycles.
  This implies that, for $(i',\chi)\in\mathrm{supp}(X)$ with $\chi=\sum_{l=1}^n \kappa_l e_{\alpha_l}$, we have $|\kappa_l|\leq K$.
